---
title: "Laplace Direct MIRT Comparison"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, message=FALSE, warning=FALSE}
library(bigIRT)
library(data.table)
library(ggplot2)
library(gridExtra)

if(!requireNamespace("mirt", quietly = TRUE)){
  stop("This notebook requires the 'mirt' package.")
}

set.seed(20260323)
theme_set(theme_minimal(base_size = 12))
```

## Goal

This notebook compares four estimators on sparse multidimensional 2PL data:

- `bigIRT` JML
- `bigIRT` `laplace_direct`
- `mirt` EM
- `mirt` MHRM

The scenarios cover:

1. 1-factor
2. 3-factor
3. 5-factor

The notebook also:

- simulates correlated ability factors when `K > 1`,
- checks whether `bigIRT`'s multivariate normalization makes direct parameter comparison simple,
- avoids truth-based ability alignment,
- uses only loading-based sign/permutation alignment for evaluation,
- reports recovery of latent correlations as well as item and ability parameters.

## Helpers

```{r helpers}
make_ability_corr <- function(K){
  if(K <= 1L) return(matrix(1, 1, 1))
  M <- outer(seq_len(K), seq_len(K), function(i, j) 0.45^(abs(i - j)))
  diag(M) <- 1
  M
}

scenario_settings <- function(K){
  list(
    Nsubs = 700,
    Nitems = 12L * K,
    Nscales = K,
    NitemsAnswered = rep(max(6L, ceiling(0.8 * K) + 6L), K),
    AbilityCorr = make_ability_corr(K)
  )
}

make_loading_mask <- function(A_true){
  mask <- matrix(0, nrow = nrow(A_true), ncol = ncol(A_true))
  mask[abs(A_true) > 1e-8] <- NA_real_
  rownames(mask) <- as.character(seq_len(nrow(mask)))
  colnames(mask) <- as.character(seq_len(ncol(mask)))
  mask
}

make_mirt_model <- function(A_true){
  item_names <- paste0("Item", seq_len(nrow(A_true)))
  lines <- character(ncol(A_true))
  for(k in seq_len(ncol(A_true))){
    active <- which(abs(A_true[, k]) > 1e-8)
    lines[k] <- paste0("F", k, " = ", paste(item_names[active], collapse = ","))
  }
  if(ncol(A_true) > 1L){
    corr_lines <- character()
    idx <- 1L
    for(i in seq_len(ncol(A_true) - 1L)){
      for(j in (i + 1L):ncol(A_true)){
        corr_lines[idx] <- paste0("COV = F", i, "*F", j)
        idx <- idx + 1L
      }
    }
    lines <- c(lines, corr_lines)
  }
  paste(lines, collapse = "\n")
}

extract_bigirt_state <- function(fit){
  list(
    A = as.matrix(fit$pars$A),
    B = as.numeric(fit$pars$B),
    Ability = as.matrix(fit$pars$Ability),
    corr = if(!is.null(fit$abilityPrior$corr)) as.matrix(fit$abilityPrior$corr) else cor(fit$pars$Ability)
  )
}

extract_mirt_state <- function(fit_obj, sim){
  coef_s <- mirt::coef(fit_obj, simplify = TRUE)
  items <- as.data.frame(coef_s$items)
  a_cols <- grep("^a[0-9]+$", names(items), value = TRUE)
  d_col <- if("d" %in% names(items)) "d" else grep("^d", names(items), value = TRUE)[1]
  A <- as.matrix(items[, a_cols, drop = FALSE])
  B <- -as.numeric(items[[d_col]])
  fs <- as.matrix(mirt::fscores(fit_obj, method = "EAP", full.scores = TRUE))
  list(
    A = A,
    B = B,
    Ability = fs,
    corr = cov2cor(coef_s$cov)
  )
}

signed_permutations <- function(K){
  permute_idx <- function(x){
    if(length(x) == 1L) return(matrix(x, nrow = 1L))
    out <- list()
    idx <- 1L
    for(i in seq_along(x)){
      rest <- x[-i]
      tail_perm <- permute_idx(rest)
      out[[idx]] <- cbind(x[i], tail_perm)
      idx <- idx + 1L
    }
    do.call(rbind, out)
  }
  perms <- permute_idx(seq_len(K))
  signs <- as.matrix(expand.grid(rep(list(c(-1, 1)), K)))
  out <- vector("list", nrow(perms) * nrow(signs))
  idx <- 1L
  for(i in seq_len(nrow(perms))){
    P <- diag(K)[, perms[i, ], drop = FALSE]
    for(j in seq_len(nrow(signs))){
      out[[idx]] <- P %*% diag(as.numeric(signs[j, ]), K)
      idx <- idx + 1L
    }
  }
  out
}

align_by_loadings <- function(A_est, Ability_est, corr_est, A_true){
  K <- ncol(A_true)
  if(K == 1L){
    c1 <- suppressWarnings(cor(A_est[,1], A_true[,1], use = "pairwise.complete.obs"))
    if(!is.finite(c1)) c1 <- sum(A_est[,1] * A_true[,1], na.rm = TRUE)
    s <- if(is.finite(c1) && c1 < 0) -1 else 1
    P <- matrix(s, 1, 1)
    return(list(
      A = A_est %*% P,
      Ability = Ability_est %*% P,
      corr = P %*% corr_est %*% P,
      transform = P
    ))
  }
  candidates <- signed_permutations(K)
  score_alignment <- function(Acand, Atrue){
    C <- suppressWarnings(cor(Acand, Atrue, use = "pairwise.complete.obs"))
    if(all(!is.finite(C))){
      C <- matrix(0, ncol(Acand), ncol(Atrue))
      for(i in seq_len(ncol(Acand))){
        xi <- Acand[, i]
        sxi <- sqrt(sum(xi^2, na.rm = TRUE))
        for(j in seq_len(ncol(Atrue))){
          yj <- Atrue[, j]
          syj <- sqrt(sum(yj^2, na.rm = TRUE))
          if(is.finite(sxi) && is.finite(syj) && sxi > 0 && syj > 0){
            C[i, j] <- sum(xi * yj, na.rm = TRUE) / (sxi * syj)
          }
        }
      }
    }
    C[!is.finite(C)] <- 0
    sum(abs(diag(C)))
  }
  scores <- vapply(candidates, function(P){
    score_alignment(A_est %*% P, A_true)
  }, numeric(1))
  if(length(scores) == 0L || all(!is.finite(scores))){
    P <- diag(K)
  } else {
    P <- candidates[[which.max(scores)]]
  }
  list(
    A = A_est %*% P,
    Ability = Ability_est %*% P,
    corr = P %*% corr_est %*% P,
    transform = P
  )
}

loading_axis_diagnostic <- function(A_est, A_true){
  C <- suppressWarnings(cor(A_est, A_true))
  data.table(
    est_factor = rep(seq_len(ncol(A_est)), times = ncol(A_true)),
    true_factor = rep(seq_len(ncol(A_true)), each = ncol(A_est)),
    corr = as.numeric(C)
  )
}

rmse <- function(x, y) sqrt(mean((x - y)^2, na.rm = TRUE))

run_bigirt_fit <- function(sim, loading_mask, method){
  t0 <- proc.time()[3]
  fit <- fitIRT(
    sim$dat,
    pl = 2,
    cores = 1,
    priors = TRUE,
    ebayes = FALSE,
    loadings = loading_mask,
    marginalApprox = if(method == "JML") "none" else "laplace_direct",
    estimateAbilityCorr = identical(method, "LaplaceDirect") && ncol(loading_mask) > 1L,
    laplaceCorrParam = "stan_corsqrt",
    dropPerfectScores = FALSE,
    normalise = TRUE,
    verbose = 0,
    plot = FALSE
  )
  list(fit = fit, elapsed = proc.time()[3] - t0)
}

run_mirt_fit <- function(sim, model_syntax, method){
  wide <- dcast(copy(sim$dat), id ~ Item, value.var = "score", fill = NA)
  setorder(wide, id)
  resp <- as.matrix(wide[, -1])
  colnames(resp) <- paste0("Item", seq_len(ncol(resp)))
  t0 <- proc.time()[3]
  fit <- mirt::mirt(
    data = resp,
    model = mirt::mirt.model(model_syntax),
    itemtype = "2PL",
    method = method,
    verbose = FALSE,
    technical = list(NCYCLES = 500L)
  )
  list(fit = fit, elapsed = proc.time()[3] - t0)
}

evaluate_method <- function(method_name, state, truth){
  aligned <- align_by_loadings(
    A_est = state$A,
    Ability_est = state$Ability,
    corr_est = state$corr,
    A_true = truth$A
  )
  list(
    metrics = data.table(
      method = method_name,
      loading_rmse = rmse(aligned$A, truth$A),
      B_rmse = rmse(state$B, truth$B),
      ability_rmse = rmse(aligned$Ability, truth$Ability),
      corr_rmse = rmse(aligned$corr, truth$corr)
    ),
    aligned = aligned,
    axis_diag = loading_axis_diagnostic(state$A, truth$A)
  )
}

make_loading_plot <- function(dt, title){
  ggplot(dt, aes(true, est, color = factor)) +
    geom_point(alpha = 0.55, size = 1.4) +
    geom_abline(slope = 1, intercept = 0, linetype = 2) +
    facet_wrap(~ factor, scales = "free") +
    labs(title = title, x = "True loading", y = "Estimated loading", color = "Factor")
}

make_corr_heatmap <- function(corr_mat, title){
  dt <- as.data.table(as.table(corr_mat))
  setnames(dt, c("f1", "f2", "corr"))
  ggplot(dt, aes(f1, f2, fill = corr)) +
    geom_tile() +
    geom_text(aes(label = sprintf("%.2f", corr)), size = 3) +
    scale_fill_gradient2(limits = c(-1, 1)) +
    coord_equal() +
    labs(title = title, x = NULL, y = NULL)
}

make_axis_plot <- function(dt, title){
  ggplot(dt, aes(factor(true_factor), factor(est_factor), fill = corr)) +
    geom_tile() +
    geom_text(aes(label = sprintf("%.2f", corr)), size = 3) +
    scale_fill_gradient2(limits = c(-1, 1)) +
    labs(title = title, x = "True factor", y = "Estimated factor")
}
```

## Scenario Runner

```{r runner}
run_scenario <- function(K){
  cfg <- scenario_settings(K)
  sim <- do.call(simIRT, c(cfg, list(
    mirt = K > 1L,
    AMean = 1.0,
    ASD = 0.20,
    loadingSparsity = 0.25,
    crossLoadingSD = 0.20,
    BMean = 0,
    BSD = 0.9,
    logitCMean = -20,
    logitCSD = 0,
    normalise = FALSE
  )))

  truth <- list(
    A = sim$A,
    B = sim$B,
    Ability = sim$Ability,
    corr = cor(sim$Ability)
  )
  loading_mask <- make_loading_mask(truth$A)
  mirt_model <- make_mirt_model(truth$A)

  fit_jml <- run_bigirt_fit(sim, loading_mask, "JML")
  fit_laplace <- run_bigirt_fit(sim, loading_mask, "LaplaceDirect")
  fit_mirt_em <- run_mirt_fit(sim, mirt_model, "EM")
  fit_mirt_mhrm <- run_mirt_fit(sim, mirt_model, "MHRM")

  states <- list(
    JML = extract_bigirt_state(fit_jml$fit),
    LaplaceDirect = extract_bigirt_state(fit_laplace$fit),
    mirt_EM = extract_mirt_state(fit_mirt_em$fit, sim),
    mirt_MHRM = extract_mirt_state(fit_mirt_mhrm$fit, sim)
  )

  evals <- lapply(names(states), function(nm) evaluate_method(nm, states[[nm]], truth))
  names(evals) <- names(states)

  metrics <- rbindlist(Map(function(nm, ev, elapsed){
    cbind(data.table(
      scenario = paste0(K, " factor"),
      elapsed_sec = elapsed
    ), ev$metrics)
  },
  names(evals),
  evals,
  c(fit_jml$elapsed, fit_laplace$elapsed, fit_mirt_em$elapsed, fit_mirt_mhrm$elapsed)))

  plots <- list(
    loading = Map(function(nm, ev){
      A_long <- data.table(
        true = as.numeric(truth$A),
        est = as.numeric(ev$aligned$A),
        factor = rep(seq_len(K), each = nrow(truth$A))
      )
      make_loading_plot(A_long, paste0(paste0(K, "-factor "), nm, ": loadings"))
    }, names(evals), evals),
    corr = Map(function(nm, ev){
      make_corr_heatmap(ev$aligned$corr, paste0(paste0(K, "-factor "), nm, ": recovered corr"))
    }, names(evals), evals),
    axis = Map(function(nm, ev){
      make_axis_plot(ev$axis_diag, paste0(paste0(K, "-factor "), nm, ": axis match before alignment"))
    }, names(evals), evals)
  )

  list(
    sim = sim,
    truth = truth,
    metrics = metrics,
    fits = list(JML = fit_jml$fit, LaplaceDirect = fit_laplace$fit,
      mirt_EM = fit_mirt_em$fit, mirt_MHRM = fit_mirt_mhrm$fit),
    states = states,
    evals = evals,
    plots = plots
  )
}
```

## Run 1-, 3-, and 5-factor scenarios

```{r run-all, cache=TRUE}
results <- lapply(c(1L, 3L, 5L), run_scenario)
names(results) <- c("1 factor", "3 factor", "5 factor")

all_metrics <- rbindlist(lapply(results, `[[`, "metrics"))
all_metrics[]
```

## Summary Table

```{r summary-table}
summary_table <- copy(all_metrics)[order(scenario, method)]
summary_table[]
```

## Timing Comparison

```{r timing-plot, fig.height=4.5, fig.width=8}
ggplot(all_metrics, aes(method, elapsed_sec, fill = method)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ scenario, scales = "free_y") +
  labs(title = "Elapsed time", x = NULL, y = "Seconds") +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

## Recovery Comparison

```{r recovery-plot, fig.height=7, fig.width=10}
metrics_long <- melt(
  all_metrics,
  id.vars = c("scenario", "method", "elapsed_sec"),
  measure.vars = c("loading_rmse", "B_rmse", "ability_rmse", "corr_rmse"),
  variable.name = "metric",
  value.name = "value"
)

ggplot(metrics_long, aes(method, value, fill = method)) +
  geom_col(show.legend = FALSE) +
  facet_grid(metric ~ scenario, scales = "free_y") +
  labs(title = "Recovery diagnostics", x = NULL, y = "RMSE") +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

## Normalization / Axis-Matching Check

These plots use the fitted normalized loadings before any sign/permutation alignment.  
If multivariate normalization is behaving well, the diagonal tiles should dominate.

```{r axis-check, fig.height=10, fig.width=10}
for(res_name in names(results)){
  cat("###", res_name, "\n")
  gridExtra::grid.arrange(grobs = results[[res_name]]$plots$axis, ncol = 2)
}
```

## Loading Recovery Plots

```{r loading-plots, fig.height=12, fig.width=10}
for(res_name in names(results)){
  cat("###", res_name, "\n")
  gridExtra::grid.arrange(grobs = results[[res_name]]$plots$loading, ncol = 2)
}
```

## Correlation Recovery Plots

```{r corr-plots, fig.height=12, fig.width=10}
for(res_name in names(results)){
  cat("###", res_name, "\n")
  truth_plot <- make_corr_heatmap(results[[res_name]]$truth$corr, paste0(res_name, ": true corr"))
  gridExtra::grid.arrange(grobs = c(list(truth_plot), results[[res_name]]$plots$corr), ncol = 2)
}
```

## Notes

- `bigIRT` fits are run with `normalise = TRUE`, so comparisons are on the fitted standardized metric rather than on an affine alignment to true abilities.
- The evaluation still applies loading-based sign/permutation alignment, because factor sign is arbitrary even under confirmatory identification and `mirt` can still return a flipped orientation.
- No truth-based ability alignment is used.
- For `K > 1`, correlation recovery is evaluated after the same loading-based alignment used for factor signs/order.
