---
title: "3-Factor MIRT: bigIRT (JML, Sampled) vs mirt"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, message=FALSE, warning=FALSE}
library(bigIRT)
library(data.table)
library(ggplot2)

set.seed(20260319)
```

## Goal

This standalone notebook demonstrates:

1. how to simulate **3-factor MIRT** data with cross-loadings using `simIRT(mirt=TRUE)`,
2. how to fit MIRT models with unknown loadings using `fitIRT(loadings=...)`,
3. how sampled-ability optimization compares to baseline JML and the `mirt` package after a rotation/scale alignment step,
4. where the full true and estimated loading matrices can be inspected directly.

## 1) Simulate 3-factor MIRT data

`simIRT(mirt=TRUE)` treats `Nitems` as total items and creates a full `A[item, factor]` loading matrix.

```{r simulate}
Nsubs <- 700
Nitems <- 36
Nscales <- 3

sim <- simIRT(
  Nsubs = Nsubs,
  Nitems = Nitems,
  Nscales = Nscales,
  NitemsAnswered = c(10, 10, 10),
  mirt = TRUE,
  AMean = 1.0,
  ASD = 0.20,
  loadingSparsity = 0.25,
  crossLoadingSD = 0.20,
  BMean = 0,
  BSD = 0.9,
  logitCMean = -20,
  logitCSD = 0
)

str(list(
  Ability_dim = dim(sim$Ability),
  A_dim = dim(sim$A),
  B_len = length(sim$B),
  dat_dim = dim(sim$dat)
))
```

Inspect true loading sparsity:

```{r loading-structure}
A_true <- sim$A
nonzero_per_item <- rowSums(abs(A_true) > 1e-8)
table(nonzero_per_item)
```

## 2) Build unknown-loading fit input

We estimate unknown loading values while keeping an interpretable sparse structure:

- entries that are truly nonzero are set to `NA` (free),
- entries that are truly zero are fixed at `0`.

This is "unknown loadings" in the sense that magnitudes are estimated, not fixed.

```{r fit-input}
loading_mask <- matrix(0, nrow = nrow(A_true), ncol = ncol(A_true))
loading_mask[abs(A_true) > 1e-8] <- NA_real_
rownames(loading_mask) <- as.character(seq_len(nrow(loading_mask)))
colnames(loading_mask) <- as.character(seq_len(ncol(loading_mask)))

summary(as.vector(is.na(loading_mask)))
```

## 3) Fit baseline JML and sampled-ability MIRT

```{r fit-models}
fit_jml_time <- system.time({
  fit_jml <- fitIRT(
    sim$dat,
    pl = 2,
    cores = 1,
    # iter = 40,
    priors = TRUE,
    ebayes = FALSE,
    loadings = loading_mask,
    dropPerfectScores = FALSE,
    normalise = FALSE,
    verbose = 0,
    plot = FALSE
  )
})

fit_samp_time <- system.time({
  fit_samp <- fitIRT(
    sim$dat,
    pl = 2,
    cores = 1,
    # iter = 30,
    priors = TRUE,
    ebayes = FALSE,
    loadings = loading_mask,
    sampledAbilityStep = TRUE,
    sampledAbilityOuterIter = 40,
    noptimsteps = 10,
    # sampledAbilitySigmaScale = 0.35,
    sampledAbilityDiagnostics = F,
    sampledAbilityPlot = T,#noptimgradtol = 1e-4,
    dropPerfectScores = FALSE,
    normalise = FALSE,
    verbose = 0,
    plot = FALSE
  )
})

fit_jml_time
fit_samp_time
```

Sampled-ability step timings (seconds) for each outer iteration:

```{r sampled-timing}
sampled_timing <- as.data.table(fit_samp$sampledAbilityTiming)
if(is.null(sampled_timing) || nrow(sampled_timing) == 0){
  stop("No sampledAbilityTiming found on fit object.")
}

timing_outer <- sampled_timing[stage == "outer",
  .(
    outerIter,
    person_point_estimate_optim_sec = round(person_opt_sec, 4),
    person_target_evals = person_target_evals,
    posterior_summary_sec = round(posterior_sec, 4),
    item_parameter_optim_sec = round(item_opt_sec, 4),
    item_target_evals = item_target_evals,
    total_outer_sec = round(total_sec, 4)
  )
]
timing_outer

timing_totals <- sampled_timing[stage == "outer",
  .(
    person_point_estimate_optim_sec = sum(person_opt_sec, na.rm = TRUE),
    person_target_evals = sum(person_target_evals, na.rm = TRUE),
    posterior_summary_sec = sum(posterior_sec, na.rm = TRUE),
    item_parameter_optim_sec = sum(item_opt_sec, na.rm = TRUE),
    item_target_evals = sum(item_target_evals, na.rm = TRUE),
    total_outer_sec = sum(total_sec, na.rm = TRUE)
  )
]
timing_totals
```

## 4) Fit `mirt` with a comparable 3-factor exploratory setup

```{r fit-mirt}
if(!requireNamespace("mirt", quietly = TRUE)){
  stop("This notebook section requires the 'mirt' package.")
}

wide <- dcast(sim$dat, id ~ Item, value.var = "score", fill = NA)
setorder(wide, id)
resp_mat <- as.matrix(wide[, -1])

fit_mirt_time <- system.time({
  fit_mirt <- mirt::mirt(
    data = resp_mat,
    model = Nscales,
    itemtype = "2PL",
    method = "MHRM",
    verbose = FALSE,
    technical = list(NCYCLES = 500L)
  )
})

fit_mirt_time
```

## 5) Alignment for sensible MIRT comparisons

In MIRT, latent solutions can differ by orthogonal rotation and factor scaling.
For apples-to-apples comparisons, align estimated abilities/loadings to the true latent frame.

```{r alignment-functions}
extract_ordered_A <- function(fit, Nitems, Nscales){
  A <- as.matrix(fit$pars$A)
  idx <- as.integer(rownames(A))
  if(any(is.na(idx))) idx <- seq_len(nrow(A))
  A[idx, seq_len(Nscales), drop = FALSE]
}

extract_ordered_theta <- function(fit, Nsubs, Nscales){
  th <- as.matrix(fit$pars$Ability)
  idx <- as.integer(rownames(th))
  if(any(is.na(idx))) idx <- seq_len(nrow(th))
  th[idx, seq_len(Nscales), drop = FALSE]
}

align_mirt_solution <- function(theta_est, A_est, theta_true, eps = 1e-8){
  mu_est <- colMeans(theta_est)
  mu_true <- colMeans(theta_true)
  th_est_c <- sweep(theta_est, 2, mu_est, "-")
  th_true_c <- sweep(theta_true, 2, mu_true, "-")

  sd_est <- pmax(apply(th_est_c, 2, sd), eps)
  sd_true <- pmax(apply(th_true_c, 2, sd), eps)

  z_est <- th_est_c %*% diag(1 / sd_est, ncol(theta_est))
  z_true <- th_true_c %*% diag(1 / sd_true, ncol(theta_true))

  M <- t(z_est) %*% z_true
  s <- svd(M)
  Q <- s$u %*% t(s$v)

  Tmat <- diag(1 / sd_est, ncol(theta_est)) %*% Q %*% diag(sd_true, ncol(theta_true))
  theta_aligned <- th_est_c %*% Tmat + matrix(rep(mu_true, each = nrow(theta_est)), nrow(theta_est))
  A_aligned <- A_est %*% solve(Tmat)

  list(theta = theta_aligned, A = A_aligned, Q = Q, transform = Tmat)
}

calc_observed_nll <- function(fit){
  p_obs <- pmin(pmax(bigIRT:::checkP(fit), 1e-10), 1 - 1e-10)
  -mean(log(p_obs))
}
```

## 6) Compute performance metrics

```{r metrics}
A_jml <- extract_ordered_A(fit_jml, Nitems, Nscales)
A_samp <- extract_ordered_A(fit_samp, Nitems, Nscales)
th_jml <- extract_ordered_theta(fit_jml, Nsubs, Nscales)
th_samp <- extract_ordered_theta(fit_samp, Nsubs, Nscales)

coef_items <- as.data.frame(mirt::coef(fit_mirt, simplify = TRUE)$items)
a_cols <- grep("^a[0-9]+$", colnames(coef_items), value = TRUE)
A_mirt_raw <- as.matrix(coef_items[, a_cols, drop = FALSE])
item_order <- as.integer(rownames(A_mirt_raw))
if(any(is.na(item_order))) item_order <- seq_len(nrow(A_mirt_raw))
A_mirt <- A_mirt_raw[order(item_order), , drop = FALSE]
th_mirt <- as.matrix(mirt::fscores(fit_mirt, method = "EAP", full.scores = TRUE))

align_jml <- align_mirt_solution(th_jml, A_jml, sim$Ability)
align_samp <- align_mirt_solution(th_samp, A_samp, sim$Ability)
align_mirt <- align_mirt_solution(th_mirt, A_mirt, sim$Ability)

free_idx <- which(is.na(loading_mask), arr.ind = TRUE)

rmse <- function(x, y) sqrt(mean((x - y)^2))

ability_rmse_jml <- rmse(align_jml$theta, sim$Ability)
ability_rmse_samp <- rmse(align_samp$theta, sim$Ability)
ability_rmse_mirt <- rmse(align_mirt$theta, sim$Ability)

loading_rmse_jml <- rmse(
  align_jml$A[free_idx],
  A_true[free_idx]
)
loading_rmse_samp <- rmse(
  align_samp$A[free_idx],
  A_true[free_idx]
)
loading_rmse_mirt <- rmse(
  align_mirt$A[free_idx],
  A_true[free_idx]
)

loading_rmse_full_jml <- rmse(align_jml$A, A_true)
loading_rmse_full_samp <- rmse(align_samp$A, A_true)
loading_rmse_full_mirt <- rmse(align_mirt$A, A_true)

nll_jml <- calc_observed_nll(fit_jml)
nll_samp <- calc_observed_nll(fit_samp)

eta_mirt <- th_mirt %*% t(A_mirt) + matrix(coef_items[order(item_order), "d"], nrow = nrow(th_mirt), ncol = nrow(A_mirt), byrow = TRUE)
p_mirt <- pmin(pmax(plogis(eta_mirt), 1e-10), 1 - 1e-10)
nll_mirt <- -mean(log(ifelse(sim$dat$score == 1, p_mirt[cbind(sim$dat$id, sim$dat$Item)], 1 - p_mirt[cbind(sim$dat$id, sim$dat$Item)])))

perf <- data.table(
  method = c("bigIRT_JML", "bigIRT_SampledAbility", "mirt_MHRM"),
  elapsed_sec = c(unname(fit_jml_time["elapsed"]), unname(fit_samp_time["elapsed"]), unname(fit_mirt_time["elapsed"])),
  observed_nll = c(nll_jml, nll_samp, nll_mirt),
  ability_rmse = c(ability_rmse_jml, ability_rmse_samp, ability_rmse_mirt),
  loading_rmse_free = c(loading_rmse_jml, loading_rmse_samp, loading_rmse_mirt),
  loading_rmse_full = c(loading_rmse_full_jml, loading_rmse_full_samp, loading_rmse_full_mirt)
)
perf
```

## 7) Full true and estimated loading matrices

```{r full-loading-matrices}
print(round(A_true, 3))
print(round(align_jml$A, 3))
print(round(align_samp$A, 3))
print(round(align_mirt$A, 3))
```

The matrices above are, in order:

- true loadings,
- bigIRT JML aligned estimates,
- bigIRT sampled-ability aligned estimates,
- mirt aligned estimates.

### Loading heatmaps

```{r loading-heatmaps}
to_heatmap_dt <- function(M, method){
  M <- as.matrix(M)
  data.table(
    item = rep(seq_len(nrow(M)), times = ncol(M)),
    factor = rep(seq_len(ncol(M)), each = nrow(M)),
    loading = as.numeric(M),
    method = method
  )
}

hm_dt <- rbindlist(list(
  to_heatmap_dt(A_true, "True"),
  to_heatmap_dt(align_jml$A, "bigIRT_JML"),
  to_heatmap_dt(align_samp$A, "bigIRT_SampledAbility"),
  to_heatmap_dt(align_mirt$A, "mirt_MHRM")
))

ggplot(hm_dt, aes(x = factor, y = item, fill = loading)) +
  geom_tile() +
  scale_fill_gradient2(low = "navy", mid = "white", high = "firebrick", midpoint = 0, limits = c(quantile(hm_dt$loading,c(.05, .95)))) +
  facet_wrap(~ method, nrow = 1) +
  scale_x_continuous(breaks = seq_len(Nscales)) +
  theme_bw() +
  labs(
    title = "Full loading matrices as heatmaps",
    x = "Latent factor",
    y = "Item"
  )
```

Interpretation:

- Brighter red/blue means larger positive/negative loadings.
- A good estimate reproduces the sparse pattern and dominant-factor structure of the true panel.
- Differences between methods are easiest to spot here when scatterplots look similar.

## 8) Visual checks

### Ability recovery

```{r ability-plot}
plot_dt <- rbind(
  data.table(
    method = "JML",
    true = as.vector(sim$Ability),
    est = as.vector(align_jml$theta)
  ),
  data.table(
    method = "SampledAbility",
    true = as.vector(sim$Ability),
    est = as.vector(align_samp$theta)
  ),
  data.table(
    method = "mirt",
    true = as.vector(sim$Ability),
    est = as.vector(align_mirt$theta)
  )
)

ggplot(plot_dt, aes(x = true, y = est)) +
  geom_point(alpha = 0.15) +
  geom_abline(intercept = 0, slope = 1, linetype = 2, color = "red") +
  facet_wrap(~ method, nrow = 1) +
  theme_bw() +
  labs(title = "Ability recovery after MIRT alignment", x = "True ability", y = "Estimated ability (aligned)")
```

### Loading recovery (free entries only)

```{r loading-plot}
ld_dt <- rbind(
  data.table(
    method = "JML",
    true = A_true[free_idx],
    est = align_jml$A[free_idx]
  ),
  data.table(
    method = "SampledAbility",
    true = A_true[free_idx],
    est = align_samp$A[free_idx]
  ),
  data.table(
    method = "mirt",
    true = A_true[free_idx],
    est = align_mirt$A[free_idx]
  )
)

ggplot(ld_dt, aes(x = true, y = est)) +
  geom_point(alpha = 0.2) +
  geom_abline(intercept = 0, slope = 1, linetype = 2, color = "red") +
  facet_wrap(~ method, nrow = 1) +
  theme_bw() +
  labs(title = "Loading recovery on free MIRT entries", x = "True loading", y = "Estimated loading (aligned)")
```

## 9) Sampled-ability diagnostics and interpretation

```{r sampled-diagnostics-check}
diagdf <- as.data.table(fit_samp$sampledAbilityDiagnostics)
if(is.null(diagdf) || nrow(diagdf) == 0){
  stop("No sampled-ability diagnostics found. Refit with sampledAbilityDiagnostics=TRUE.")
}

diagdf[, iter_step := seq_len(.N)]
diagdf[]

timing_diag <- as.data.table(fit_samp$sampledAbilityTiming)[stage == "outer"]
timing_diag[]
```

### Accepted vs rejected outer iterations

```{r sampled-diagnostics-acceptance-summary}
timing_diag[, .(
  outerIter,
  accepted,
  rejected,
  reject_reason,
  sigmaScaleUsed,
  meanPosteriorSD_ratio,
  maxPosteriorSD_ratio,
  combinedGradNorm
)]
```

### Combined gradient norm over sampled steps

```{r sampled-diagnostics-grad}
ggplot(diagdf[stage %in% c("item", "person")],
  aes(x = iter_step, y = combinedGradNorm, color = stage)) +
  geom_line(linewidth = 0.6) +
  scale_y_log10() +
  theme_bw() +
  labs(
    title = "Sampled-ability convergence signal",
    x = "Sampled update step",
    y = "Combined gradient norm (log10)"
  )
```

### Log-likelihood trajectory (stabilized scale)

```{r sampled-diagnostics-loglik}
loglik_dt <- copy(diagdf[stage %in% c("item", "person"), .(iter_step, stage, logLik)])
if(!"logLik" %in% colnames(loglik_dt) || all(!is.finite(loglik_dt$logLik))){
  warning("logLik not available in sampled diagnostics for this fit.")
} else {
  min_loglik <- min(loglik_dt$logLik[is.finite(loglik_dt$logLik)])
  loglik_dt[, logLik_shift := log1p(pmax(logLik - min_loglik, 0))]

  ggplot(loglik_dt, aes(x = iter_step, y = logLik_shift, color = stage)) +
    geom_line(linewidth = 0.7) +
    theme_bw() +
    labs(
      title = "Sampled objective trajectory: log(1 + logLik - min(logLik))",
      x = "Sampled update step",
      y = "log(1 + logLik - min(logLik))"
    )
}
```

### Acceptance / rejection path

```{r sampled-diagnostics-acceptance}
ggplot(timing_diag, aes(x = outerIter, y = as.numeric(accepted), color = reject_reason)) +
  geom_point(size = 2) +
  geom_line(linewidth = 0.5) +
  scale_y_continuous(breaks = c(0, 1), labels = c("reject", "accept")) +
  theme_bw() +
  labs(
    title = "Accepted vs rejected sampled outer iterations",
    x = "Outer iteration",
    y = NULL,
    color = "Reject reason"
  )
```

### Sigma-scale trajectory

```{r sampled-diagnostics-sigmascale}
ggplot(timing_diag, aes(x = outerIter, y = sigmaScaleUsed, color = accepted)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2) +
  theme_bw() +
  labs(
    title = "Adaptive sigma scale during sampled updates",
    x = "Outer iteration",
    y = "Sigma scale"
  )
```

### Item vs person step movement

```{r sampled-diagnostics-rms}
move_dt <- melt(
  diagdf[stage %in% c("item", "person"),
    .(iter_step, stage, itemStepRms, personStepRms)],
  id.vars = c("iter_step", "stage"),
  variable.name = "movement_type",
  value.name = "rms"
)

ggplot(move_dt[is.finite(rms)], aes(x = iter_step, y = rms, color = movement_type)) +
  geom_line(linewidth = 0.6) +
  scale_y_log10() +
  facet_wrap(~ stage, ncol = 1) +
  theme_bw() +
  labs(
    title = "Parameter movement by alternating sub-step",
    x = "Sampled update step",
    y = "RMS parameter movement (log10)"
  )
```

### Posterior SD behavior during sampled updates

```{r sampled-diagnostics-posterior-sd}
ggplot(diagdf[stage %in% c("item", "person")], aes(x = iter_step)) +
  geom_line(aes(y = meanPosteriorSD, color = "meanPosteriorSD"), linewidth = 0.7) +
  geom_line(aes(y = maxPosteriorSD, color = "maxPosteriorSD"), linewidth = 0.7, linetype = 2) +
  theme_bw() +
  labs(
    title = "Posterior uncertainty summary during sampled updates",
    x = "Sampled update step",
    y = "Posterior SD",
    color = NULL
  )
```

### Posterior SD growth ratios

```{r sampled-diagnostics-posterior-ratio}
ggplot(timing_diag, aes(x = outerIter)) +
  geom_line(aes(y = meanPosteriorSD_ratio, color = "meanPosteriorSD_ratio"), linewidth = 0.7) +
  geom_line(aes(y = maxPosteriorSD_ratio, color = "maxPosteriorSD_ratio"), linewidth = 0.7, linetype = 2) +
  geom_hline(yintercept = 1.2, linetype = 2, color = "grey50") +
  geom_hline(yintercept = 1.35, linetype = 3, color = "grey50") +
  theme_bw() +
  labs(
    title = "Posterior spread ratios by outer iteration",
    x = "Outer iteration",
    y = "Ratio to previous accepted posterior spread",
    color = NULL
  )
```

### Raw log-likelihood by sub-step (stabilized scale)

```{r sampled-diagnostics-objective}
obj_raw <- copy(diagdf[stage %in% c("person", "item"), .(iter_step, outerIter, stage, logLik)])
if(!"logLik" %in% colnames(obj_raw) || all(!is.finite(obj_raw$logLik))){
  warning("Raw step logLik not available in sampled diagnostics for this fit.")
} else {
  min_loglik <- min(obj_raw$logLik[is.finite(obj_raw$logLik)])
  obj_raw[, logLik_shift := log1p(pmax(logLik - min_loglik, 0))]

  ggplot(obj_raw, aes(x = iter_step, y = logLik_shift, color = stage)) +
    geom_line(linewidth = 0.7) +
    geom_point(size = 1.6) +
    theme_bw() +
    labs(
      title = "Raw sampled step objective: log(1 + logLik - min(logLik))",
      x = "Sampled update step",
      y = "log(1 + logLik - min(logLik))",
      color = NULL
    )
}
```

How to interpret these diagnostic plots:

- **Combined gradient norm** should trend downward; flattening near tolerance indicates convergence.
- **Item/person RMS movement** should generally shrink across outer iterations; spikes can indicate local re-adjustments.
- **Posterior SD curves** reflect uncertainty calibration as parameters settle; explosive growth suggests instability/jitter issues.
- **Accepted/rejected path** should contain mostly accepted iterations once the loop stabilizes.
- **Sigma-scale trajectory** should contract after rejected or unstable iterations and may expand again once the loop settles.
- **Objective deltas** should usually stay non-negative after backtracking; repeated negatives indicate acceptance-policy bugs.

## 10) Notes

- This workflow uses sparse structure with unknown loading magnitudes.
- Alignment is essential in MIRT; raw factor-wise comparisons can be misleading.
- You can stress-test with:
  - fewer responses per person (`NitemsAnswered`),
  - more cross-loadings (`loadingSparsity`),
  - weaker primary loadings (`ASD`, `AMean`).
