---
title: "3-Factor MIRT: bigIRT Laplace-EM (MML approx) vs JML vs mirt"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, message=FALSE, warning=FALSE}
library(bigIRT)
library(data.table)
library(ggplot2)

set.seed(20260320)
```

## Goal

This notebook now focuses on the `laplace_em` marginal-likelihood approximation in `bigIRT`:

1. fit the same sparse 3-factor model with `bigIRT` JML, `bigIRT` Laplace-EM, and multiple `mirt` estimators,
2. report timing and fit quality for each approach tested,
3. compare recovery quality with stronger diagnostics and plots,
4. explicitly test two latent-space normalization/alignment strategies because the earlier one can be unfavorable to `mirt`.

## 1) Simulate sparse 3-factor MIRT data

```{r simulate}
Nsubs <- 700
Nitems <- 36
Nscales <- 3

sim <- simIRT(
  Nsubs = Nsubs,
  Nitems = Nitems,
  Nscales = Nscales,
  NitemsAnswered = c(10, 10, 10),
  mirt = TRUE,
  AMean = 1.0,
  ASD = 0.20,
  loadingSparsity = 0.25,
  crossLoadingSD = 0.20,
  BMean = 0,
  BSD = 0.9,
  logitCMean = -20,
  logitCSD = 0
)

A_true <- sim$A
free_mask <- abs(A_true) > 1e-8
table(rowSums(free_mask))
```

## 2) Build unknown-loading structure

We estimate only entries that are truly nonzero (set to `NA`), while fixed zeros remain `0`.

```{r fit-input}
loading_mask <- matrix(0, nrow = nrow(A_true), ncol = ncol(A_true))
loading_mask[free_mask] <- NA_real_
rownames(loading_mask) <- as.character(seq_len(nrow(loading_mask)))
colnames(loading_mask) <- as.character(seq_len(ncol(loading_mask)))
summary(as.vector(is.na(loading_mask)))
```

## 3) Fit JML, Laplace-EM, and three mirt estimators

```{r fit-models}
fit_jml_time <- system.time({
  fit_jml <- fitIRT(
    sim$dat,
    pl = 2,
    cores = 1,
    priors = TRUE,
    ebayes = FALSE,
    loadings = loading_mask,
    dropPerfectScores = FALSE,
    normalise = FALSE,
    verbose = 0,
    plot = FALSE
  )
})

fit_laplace_time <- system.time({
  fit_laplace <- fitIRT(
    sim$dat,
    pl = 2,
    cores = 1,
    priors = TRUE,
    ebayes = FALSE,
    loadings = loading_mask,#laplacePlot = T,
    marginalApprox = "laplace_em",#noptimsteps = 20,
    dropPerfectScores = FALSE,
    normalise = FALSE,
    verbose = 0,
    plot = FALSE
  )
})

if(!requireNamespace("mirt", quietly = TRUE)){
  stop("This notebook requires the 'mirt' package.")
}

wide <- dcast(sim$dat, id ~ Item, value.var = "score", fill = NA)
setorder(wide, id)
resp_mat <- as.matrix(wide[, -1])

mirt_methods <- c("MHRM", "QMCEM", "EM")
fit_mirt <- list()
mirt_fit_status <- rbindlist(lapply(mirt_methods, function(m){
  t0 <- proc.time()
  fit_try <- tryCatch(
    mirt::mirt(
      data = resp_mat,
      model = Nscales,
      itemtype = "2PL",
      method = m,
      verbose = FALSE,
      technical = list(NCYCLES = 500L)
    ),
    error = function(e) e
  )
  elapsed <- unname((proc.time() - t0)[["elapsed"]])
  if(inherits(fit_try, "error")){
    data.table(method = m, status = "failed", elapsed_sec = elapsed, error_message = conditionMessage(fit_try))
  } else {
    fit_mirt[[m]] <<- fit_try
    data.table(method = m, status = "ok", elapsed_sec = elapsed, error_message = NA_character_)
  }
}))

if(!any(mirt_fit_status$status == "ok")){
  stop("All requested mirt estimation methods failed.")
}

fit_jml_time
fit_laplace_time
mirt_fit_status
```

## 4) Alignment / normalization strategies

The previous notebook used a z-score + orthogonal rotation alignment only.  
That can understate `mirt` when the recovered factor metric includes non-orthogonal mixing.

We compare:

- `zrot`: z-score each factor then orthogonal Procrustes rotation
- `affine`: full linear map from estimated abilities to true abilities (least squares), then induced inverse map on loadings

```{r alignment-functions}
extract_ordered_A <- function(fit, Nitems, Nscales){
  A <- as.matrix(fit$pars$A)
  idx <- as.integer(rownames(A))
  if(any(is.na(idx))) idx <- seq_len(nrow(A))
  A[idx, seq_len(Nscales), drop = FALSE]
}

extract_ordered_theta <- function(fit, Nsubs, Nscales){
  th <- as.matrix(fit$pars$Ability)
  idx <- as.integer(rownames(th))
  if(any(is.na(idx))) idx <- seq_len(nrow(th))
  th[idx, seq_len(Nscales), drop = FALSE]
}

align_zrot <- function(theta_est, A_est, theta_true, eps = 1e-8){
  mu_est <- colMeans(theta_est)
  mu_true <- colMeans(theta_true)
  th_est_c <- sweep(theta_est, 2, mu_est, "-")
  th_true_c <- sweep(theta_true, 2, mu_true, "-")

  sd_est <- pmax(apply(th_est_c, 2, sd), eps)
  sd_true <- pmax(apply(th_true_c, 2, sd), eps)

  z_est <- th_est_c %*% diag(1 / sd_est, ncol(theta_est))
  z_true <- th_true_c %*% diag(1 / sd_true, ncol(theta_true))

  M <- t(z_est) %*% z_true
  s <- svd(M)
  Q <- s$u %*% t(s$v)

  Tmat <- diag(1 / sd_est, ncol(theta_est)) %*% Q %*% diag(sd_true, ncol(theta_true))
  theta_aligned <- th_est_c %*% Tmat + matrix(rep(mu_true, each = nrow(theta_est)), nrow(theta_est))
  A_aligned <- A_est %*% solve(Tmat)

  list(theta = theta_aligned, A = A_aligned, transform = Tmat)
}

align_affine <- function(theta_est, A_est, theta_true, eps = 1e-8){
  mu_est <- colMeans(theta_est)
  mu_true <- colMeans(theta_true)
  X <- sweep(theta_est, 2, mu_est, "-")
  Y <- sweep(theta_true, 2, mu_true, "-")

  XtX <- crossprod(X) + diag(eps, ncol(X))
  Xty <- crossprod(X, Y)
  Tmat <- solve(XtX, Xty)
  invT <- tryCatch(solve(Tmat), error = function(e) qr.solve(Tmat, diag(ncol(Tmat))))

  theta_aligned <- X %*% Tmat + matrix(rep(mu_true, each = nrow(theta_est)), nrow(theta_est))
  A_aligned <- A_est %*% invT

  list(theta = theta_aligned, A = A_aligned, transform = Tmat)
}

rmse <- function(x, y) sqrt(mean((x - y)^2))

calc_observed_nll <- function(fit){
  p_obs <- pmin(pmax(bigIRT:::checkP(fit), 1e-10), 1 - 1e-10)
  -mean(log(p_obs))
}

compress_scatter <- function(dt, x_col, y_col, p = 0.95){
  x <- dt[[x_col]]
  y <- dt[[y_col]]
  bx <- as.numeric(quantile(abs(x), probs = p, na.rm = TRUE))
  by <- as.numeric(quantile(abs(y), probs = p, na.rm = TRUE))
  if(!is.finite(bx) || bx <= 0) bx <- max(abs(x), na.rm = TRUE)
  if(!is.finite(by) || by <= 0) by <- max(abs(y), na.rm = TRUE)
  dt[, x_plot := pmax(pmin(get(x_col), bx), -bx)]
  dt[, y_plot := pmax(pmin(get(y_col), by), -by)]
  dt[, outlier_flag := fifelse(abs(get(x_col)) > bx | abs(get(y_col)) > by, "clipped_outlier", "in_range")]
  dt[, `:=`(clip_x = bx, clip_y = by)]
  dt
}
```

## 5) Extract model objects and align

```{r extract-and-align}
A_jml <- extract_ordered_A(fit_jml, Nitems, Nscales)
A_laplace <- extract_ordered_A(fit_laplace, Nitems, Nscales)
th_jml <- extract_ordered_theta(fit_jml, Nsubs, Nscales)
th_laplace <- extract_ordered_theta(fit_laplace, Nsubs, Nscales)

extract_mirt_state <- function(fit_obj, method_name){
  coef_items <- as.data.frame(mirt::coef(fit_obj, simplify = TRUE)$items)
  a_cols <- grep("^a[0-9]+$", colnames(coef_items), value = TRUE)
  A_raw <- as.matrix(coef_items[, a_cols, drop = FALSE])
  item_order <- as.integer(rownames(A_raw))
  if(any(is.na(item_order))) item_order <- seq_len(nrow(A_raw))
  ord <- order(item_order)
  A <- A_raw[ord, , drop = FALSE]
  th <- as.matrix(mirt::fscores(fit_obj, method = "EAP", full.scores = TRUE))
  d <- coef_items[ord, "d"]
  eta <- th %*% t(A) + matrix(d, nrow = nrow(th), ncol = nrow(A), byrow = TRUE)
  p_full <- pmin(pmax(plogis(eta), 1e-10), 1 - 1e-10)
  p_obs <- p_full[cbind(sim$dat$id, sim$dat$Item)]
  list(
    method = paste0("mirt_", method_name),
    theta = th,
    A = A,
    p_obs = p_obs,
    observed_nll = -mean(log(ifelse(sim$dat$score == 1, p_obs, 1 - p_obs)))
  )
}

model_states <- list(
  JML = list(
    method = "JML",
    theta = th_jml,
    A = A_jml,
    p_obs = pmin(pmax(bigIRT:::checkP(fit_jml), 1e-10), 1 - 1e-10),
    observed_nll = calc_observed_nll(fit_jml)
  ),
  LaplaceEM = list(
    method = "LaplaceEM",
    theta = th_laplace,
    A = A_laplace,
    p_obs = pmin(pmax(bigIRT:::checkP(fit_laplace), 1e-10), 1 - 1e-10),
    observed_nll = calc_observed_nll(fit_laplace)
  )
)

ok_mirt <- mirt_fit_status[status == "ok", method]
for(m in ok_mirt){
  st <- extract_mirt_state(fit_mirt[[m]], m)
  model_states[[st$method]] <- st
}

alignments <- list(
  zrot = lapply(model_states, function(st) align_zrot(st$theta, st$A, sim$Ability)),
  affine = lapply(model_states, function(st) align_affine(st$theta, st$A, sim$Ability))
)

free_idx <- which(is.na(loading_mask), arr.ind = TRUE)
```

## 6) Metric table: normalization sensitivity + core comparison

```{r performance}
metric_rows <- rbindlist(lapply(names(alignments), function(aln_name){
  aln <- alignments[[aln_name]]
  rbindlist(lapply(names(aln), function(method_name){
    data.table(
      alignment = aln_name,
      method = method_name,
      ability_rmse = rmse(aln[[method_name]]$theta, sim$Ability),
      loading_rmse_free = rmse(aln[[method_name]]$A[free_idx], A_true[free_idx]),
      loading_rmse_full = rmse(aln[[method_name]]$A, A_true)
    )
  }))
}))

elapsed <- data.table(
  method = c("JML", "LaplaceEM", paste0("mirt_", ok_mirt)),
  elapsed_sec = c(
    unname(fit_jml_time["elapsed"]),
    unname(fit_laplace_time["elapsed"]),
    mirt_fit_status[match(ok_mirt, method), elapsed_sec]
  ),
  observed_nll = c(
    model_states$JML$observed_nll,
    model_states$LaplaceEM$observed_nll,
    sapply(ok_mirt, function(m) model_states[[paste0("mirt_", m)]]$observed_nll)
  )
)

perf <- merge(metric_rows, elapsed, by = "method")
setcolorder(perf, c("alignment", "method", "elapsed_sec", "observed_nll", "ability_rmse", "loading_rmse_free", "loading_rmse_full"))
perf[order(alignment, method)]
```

```{r mirt-method-timings}
mirt_fit_status[order(status, elapsed_sec)]
```

```{r all-approach-timings}
timing_all <- rbindlist(list(
  data.table(method = "JML", status = "ok", elapsed_sec = unname(fit_jml_time["elapsed"]), error_message = NA_character_),
  data.table(method = "LaplaceEM", status = "ok", elapsed_sec = unname(fit_laplace_time["elapsed"]), error_message = NA_character_),
  mirt_fit_status[, .(method = paste0("mirt_", method), status, elapsed_sec, error_message)]
), use.names = TRUE, fill = TRUE)

timing_all[order(status, elapsed_sec)]
```

Use `alignment == "affine"` as the primary comparison below, since it generally treats `mirt` more fairly when latent metrics are mixed.

```{r primary-alignment-select}
primary <- alignments$affine
```

## 7) Recovery plots (JML, Laplace-EM, and tested mirt estimators)

### Ability recovery by factor

```{r ability-by-factor}
ability_dt <- rbindlist(lapply(names(primary), function(m){
  data.table(
    id = rep(seq_len(Nsubs), times = Nscales),
    factor = rep(paste0("F", seq_len(Nscales)), each = Nsubs),
    method = m,
    true = as.vector(sim$Ability),
    est = as.vector(primary[[m]]$theta)
  )
}))

ability_dt <- compress_scatter(ability_dt, "true", "est", p = 0.95)

ggplot(ability_dt, aes(x = x_plot, y = y_plot)) +
  geom_point(aes(shape = outlier_flag, color = outlier_flag), alpha = 0.30, size = 1.0) +
  geom_abline(intercept = 0, slope = 1, linetype = 2, color = "red") +
  facet_grid(factor ~ method) +
  scale_shape_manual(values = c(in_range = 16, clipped_outlier = 17)) +
  scale_color_manual(values = c(in_range = "grey35", clipped_outlier = "darkorange3")) +
  theme_bw() +
  labs(
    title = "Ability recovery by factor (affine, 95% clipping)",
    x = "True ability (clipped at 95% |value|)",
    y = "Estimated ability (clipped at 95% |value|)",
    shape = NULL,
    color = NULL
  )
```

### Loading recovery by factor (free entries)

```{r loading-by-factor}
loading_dt <- rbindlist(lapply(names(primary), function(m){
  idx <- free_idx
  data.table(
    method = m,
    factor = paste0("F", idx[, 2]),
    true = A_true[idx],
    est = primary[[m]]$A[idx],
    abs_err = abs(primary[[m]]$A[idx] - A_true[idx])
  )
}))

loading_dt <- compress_scatter(loading_dt, "true", "est", p = 0.95)

ggplot(loading_dt, aes(x = x_plot, y = y_plot)) +
  geom_point(aes(shape = outlier_flag, color = outlier_flag), alpha = 0.35, size = 1.1) +
  geom_abline(intercept = 0, slope = 1, linetype = 2, color = "red") +
  facet_grid(factor ~ method) +
  scale_shape_manual(values = c(in_range = 16, clipped_outlier = 17)) +
  scale_color_manual(values = c(in_range = "grey35", clipped_outlier = "darkorange3")) +
  theme_bw() +
  labs(
    title = "Loading recovery on free entries (affine, 95% clipping)",
    x = "True loading (clipped at 95% |value|)",
    y = "Estimated loading (clipped at 95% |value|)",
    shape = NULL,
    color = NULL
  )
```

### Per-factor ability correlation

```{r factor-correlation}
corr_dt <- ability_dt[, .(corr = cor(true, est)), by = .(method, factor)]

ggplot(corr_dt, aes(x = factor, y = corr, fill = method)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.65) +
  coord_cartesian(ylim = c(0, 1)) +
  theme_bw() +
  labs(
    title = "Per-factor ability correlation with truth",
    x = "Factor",
    y = "Correlation"
  )

corr_dt[order(factor, -corr)]
```

### Loading absolute-error distribution

```{r loading-error-distribution}
err_cap <- as.numeric(quantile(loading_dt$abs_err, probs = 0.95, na.rm = TRUE))
if(!is.finite(err_cap) || err_cap <= 0) err_cap <- max(loading_dt$abs_err, na.rm = TRUE)
loading_dt[, abs_err_plot := pmin(abs_err, err_cap)]
loading_dt[, err_outlier := fifelse(abs_err > err_cap, "clipped_outlier", "in_range")]

ggplot(loading_dt, aes(x = method, y = abs_err_plot, fill = method)) +
  geom_violin(trim = TRUE, alpha = 0.45, color = NA) +
  geom_boxplot(width = 0.15, outlier.shape = NA) +
  geom_jitter(
    aes(color = err_outlier, shape = err_outlier),
    width = 0.12,
    height = 0,
    alpha = 0.35,
    size = 1.0
  ) +
  facet_wrap(~ factor, nrow = 1) +
  scale_shape_manual(values = c(in_range = 16, clipped_outlier = 17)) +
  scale_color_manual(values = c(in_range = "grey35", clipped_outlier = "darkorange3")) +
  theme_bw() +
  labs(
    title = "Distribution of |loading error| on free entries (95% clipping)",
    x = NULL,
    y = "Absolute error (clipped at 95% quantile)",
    color = NULL,
    shape = NULL
  )
```

### Probability calibration on observed responses

```{r calibration}
calib <- rbindlist(lapply(names(model_states), function(m){
  data.table(method = m, p = model_states[[m]]$p_obs, y = sim$dat$score)
}))

calib[, bin := cut(p, breaks = seq(0, 1, by = 0.1), include.lowest = TRUE)]
calib_dt <- calib[, .(
  mean_pred = mean(p),
  obs_rate = mean(y),
  n = .N
), by = .(method, bin)]

ggplot(calib_dt, aes(x = mean_pred, y = obs_rate, color = method)) +
  geom_abline(intercept = 0, slope = 1, linetype = 2, color = "grey50") +
  geom_point(aes(size = n), alpha = 0.85) +
  geom_line(alpha = 0.6) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  theme_bw() +
  labs(
    title = "Observed-response calibration",
    x = "Mean predicted probability",
    y = "Observed proportion correct",
    size = "Count"
  )

calib[, .(
  logloss = -mean(log(ifelse(y == 1, p, 1 - p))),
  brier = mean((y - p)^2)
), by = method][order(logloss)]
```

## 8) Loading heatmaps

```{r loading-heatmaps}
to_heatmap_dt <- function(M, method){
  M <- as.matrix(M)
  data.table(
    item = rep(seq_len(nrow(M)), times = ncol(M)),
    factor = rep(seq_len(ncol(M)), each = nrow(M)),
    loading = as.numeric(M),
    method = method
  )
}

hm_dt <- rbindlist(list(
  to_heatmap_dt(A_true, "True")
))
hm_dt <- rbind(
  hm_dt,
  rbindlist(lapply(names(primary), function(m) to_heatmap_dt(primary[[m]]$A, m)))
)

lim <- quantile(abs(hm_dt$loading), probs = 0.95, na.rm = TRUE)
ggplot(hm_dt, aes(x = factor, y = item, fill = loading)) +
  geom_tile() +
  scale_fill_gradient2(
    low = "navy",
    mid = "white",
    high = "firebrick",
    midpoint = 0,
    limits = c(-lim, lim)
  ) +
  facet_wrap(~ method, nrow = 1) +
  scale_x_continuous(breaks = seq_len(Nscales)) +
  theme_bw() +
  labs(
    title = "Loading matrices (affine-aligned estimates)",
    x = "Latent factor",
    y = "Item"
  )
```

## 9) Notes

- This version centers the comparison on `marginalApprox = "laplace_em"` as the MML approximation path.
- It tests three relevant `mirt` MIRT estimators (`MHRM`, `QMCEM`, `MCEM`) and records timing/status for each.
- Sampled-ability-specific timing/diagnostic sections were removed.
- For cross-package MIRT comparisons, alignment choice matters: check both `zrot` and `affine`, and report both when methods look unexpectedly different.
- Scatter recovery plots apply 95% absolute-value clipping and mark clipped points with a separate color/shape.
