---
title: "3-Factor MIRT: bigIRT Methods vs mirt"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, message=FALSE, warning=FALSE}
library(bigIRT)
library(data.table)
library(ggplot2)

# set.seed(20260320)
```

## Goal

This notebook compares a sparse 3-factor MIRT fit across:

1. selected `bigIRT` methods from `bigirt_methods` below,
2. selected `mirt` methods from `mirt_methods` below.

All parameter extraction and normalization now go through `compareIRTmodels()`, so
truth and fitted models are compared on one package-level normalized metric.

## 1) Simulate sparse 3-factor MIRT data

```{r simulate}
Nsubs <- 4700
Nitems <- 36
Nscales <- 3

sim <- simIRT(
  Nsubs = Nsubs,
  Nitems = Nitems,
  Nscales = Nscales,
  NitemsAnswered = c(4, 4, 4),
  mirt = TRUE,
  AMean = 1.0,
  ASD = 0.20,
  loadingSparsity = 0.25,
  crossLoadingSD = 0.20,
  BMean = 0,
  BSD = 0.9,
  logitCMean = 0,
  logitCSD = 1,
  logitDMean = 1,
  logitDSD = 1
)

A_true <- sim$A
free_mask <- abs(A_true) > 1e-8
table(rowSums(free_mask))
```

## 2) Build unknown-loading structure

We estimate only entries that are truly nonzero (set to `NA`), while fixed zeros remain `0`.

```{r fit-input}
loading_mask <- matrix(0, nrow = nrow(A_true), ncol = ncol(A_true))
loading_mask[free_mask] <- NA_real_
rownames(loading_mask) <- as.character(seq_len(nrow(loading_mask)))
colnames(loading_mask) <- as.character(seq_len(ncol(loading_mask)))
summary(as.vector(is.na(loading_mask)))
```

## 3) Fit selected bigIRT and mirt estimators

```{r fit-models}
bigirt_methods <- list(
  LaplaceDirect = list(marginalApprox = "laplace_direct")
)
# To include JML as well:
# bigirt_methods$JML <- list()

base_bigirt_args <- list(
  dat = sim$dat,
  pl = 4,
  cores = 1,
  priors = TRUE,
  ebayes = FALSE,
  loadings = loading_mask,
  dropPerfectScores = FALSE,
  normalise = FALSE,
  verbose = 0,
  plot = FALSE
)

fit_bigirt <- list()
bigirt_fit_status <- rbindlist(lapply(names(bigirt_methods), function(m){
  t0 <- proc.time()
  fit_try <- tryCatch(
    do.call(fitIRT, c(base_bigirt_args, bigirt_methods[[m]])),
    error = function(e) e
  )
  elapsed <- unname((proc.time() - t0)[["elapsed"]])
  if(inherits(fit_try, "error")){
    data.table(method = m, status = "failed", elapsed_sec = elapsed, error_message = conditionMessage(fit_try))
  } else {
    fit_bigirt[[m]] <<- fit_try
    data.table(method = m, status = "ok", elapsed_sec = elapsed, error_message = NA_character_)
  }
}))

if(!requireNamespace("mirt", quietly = TRUE)){
  stop("This notebook requires the 'mirt' package.")
}

wide <- dcast(sim$dat, id ~ Item, value.var = "score", fill = NA)
setorder(wide, id)
resp_mat <- as.matrix(wide[, -1])

mirt_methods <- c("EM") #"MHRM", 
fit_mirt <- list()
build_mirt_model_from_loadings <- function(A_true){
  lines <- vapply(seq_len(ncol(A_true)), function(k){
    active <- which(abs(A_true[, k]) > 1e-8)
    paste0("F", k, " = ", paste(active, collapse = ","))
  }, character(1))
  mirt::mirt.model(paste(lines, collapse = "\n"))
}
mirt_model <- build_mirt_model_from_loadings(A_true)
mirt_itemtype <- "4PL"
mirt_fit_status <- rbindlist(lapply(mirt_methods, function(m){
  t0 <- proc.time()
  fit_try <- tryCatch(
    mirt::mirt(
      data = resp_mat,
      model = mirt_model,
      itemtype = mirt_itemtype,
      method = m,
      verbose = TRUE, TOL=1e-2, #check this
      technical = list(NCYCLES = 500L)
    ),
    error = function(e) e
  )
  elapsed <- unname((proc.time() - t0)[["elapsed"]])
  if(inherits(fit_try, "error")){
    data.table(method = m, status = "failed", elapsed_sec = elapsed, error_message = conditionMessage(fit_try))
  } else {
    fit_mirt[[m]] <<- fit_try
    data.table(method = m, status = "ok", elapsed_sec = elapsed, error_message = NA_character_)
  }
}))

if(!any(bigirt_fit_status$status == "ok")){
  stop("All requested bigIRT estimation methods failed.")
}

if(!any(mirt_fit_status$status == "ok")){
  stop("All requested mirt estimation methods failed.")
}

bigirt_fit_status
mirt_fit_status
```

## 4) Build a common comparison object

```{r comparison-helpers}
rmse <- function(x, y) sqrt(mean((x - y)^2))

calc_observed_response_metrics <- function(model, dat){
  state <- extractMIRTpars(model)
  pcorrect_mat <- as.matrix(IRTcurve(
    A = state$A,
    B = state$B,
    C = state$C,
    D = state$D,
    theta = state$Ability,
    plot = FALSE
  ))
  pcorrect <- pcorrect_mat[cbind(dat$id, dat$Item)]
  p_obs <- dat$score * pcorrect + (1 - dat$score) * (1 - pcorrect)
  p_obs <- pmin(pmax(p_obs, 1e-10), 1 - 1e-10)
  list(
    pcorrect = pcorrect,
    p_obs = p_obs,
    observed_nll = -mean(log(p_obs))
  )
}

compress_scatter <- function(dt, x_col, y_col, p = 0.98){
  x <- dt[[x_col]]
  y <- dt[[y_col]]
  bx <- as.numeric(quantile(abs(x), probs = p, na.rm = TRUE))
  by <- as.numeric(quantile(abs(y), probs = p, na.rm = TRUE))
  if(!is.finite(bx) || bx <= 0) bx <- max(abs(x), na.rm = TRUE)
  if(!is.finite(by) || by <= 0) by <- max(abs(y), na.rm = TRUE)
  dt[, x_plot := pmax(pmin(get(x_col), bx), -bx)]
  dt[, y_plot := pmax(pmin(get(y_col), by), -by)]
  dt[, outlier_flag := fifelse(abs(get(x_col)) > bx | abs(get(y_col)) > by, "clipped_outlier", "in_range")]
  dt[, `:=`(clip_x = bx, clip_y = by)]
  dt
}

ok_bigirt <- bigirt_fit_status[status == "ok", method]
ok_mirt <- mirt_fit_status[status == "ok", method]
comparison_inputs <- c(
  list(truth = sim),
  fit_bigirt[ok_bigirt],
  setNames(fit_mirt[ok_mirt], paste0("mirt_", ok_mirt))
)

comparison <- compareIRTmodels(comparison_inputs)

model_states <- c(
  setNames(lapply(ok_bigirt, function(m){
    calc_observed_response_metrics(fit_bigirt[[m]], sim$dat)
  }), ok_bigirt),
  setNames(lapply(ok_mirt, function(m){
    calc_observed_response_metrics(fit_mirt[[m]], sim$dat)
  }), paste0("mirt_", ok_mirt))
)

item_norm <- comparison$itempars[form == "normalized"]
person_norm <- comparison$personpars[form == "normalized"]
loading_norm <- comparison$loading_matrix[form == "normalized"]
corr_raw <- comparison$ability_corr_matrix[form == "raw" & corr_type == "estimated"]

free_items <- data.table(
  item = rownames(loading_mask)[row(loading_mask)[is.na(loading_mask)]],
  factor = paste0("F", col(loading_mask)[is.na(loading_mask)])
)

loading_free <- merge(
  item_norm[parameter == "A" & source != "truth"],
  free_items,
  by = c("item", "factor")
)
```

## 5) Metric table: recovery on the normalized scale

```{r performance}
metric_rows <- rbindlist(lapply(setdiff(unique(item_norm$source), "truth"), function(method_name){
  ability_dt <- person_norm[source == method_name]
  loading_full_dt <- item_norm[source == method_name & parameter == "A"]
  loading_free_dt <- loading_free[source == method_name]
  corr_dt <- corr_raw[source == method_name]
  data.table(
    method = method_name,
    ability_rmse = rmse(ability_dt$value, ability_dt$referenceVal),
    loading_rmse_free = rmse(loading_free_dt$value, loading_free_dt$referenceVal),
    loading_rmse_full = rmse(loading_full_dt$value, loading_full_dt$referenceVal),
    corr_rmse = rmse(corr_dt$value, corr_dt$referenceVal)
  )
}))

elapsed <- data.table(
  method = c(ok_bigirt, paste0("mirt_", ok_mirt)),
  elapsed_sec = c(
    bigirt_fit_status[match(ok_bigirt, method), elapsed_sec],
    mirt_fit_status[match(ok_mirt, method), elapsed_sec]
  ),
  observed_nll = c(
    sapply(ok_bigirt, function(m) model_states[[m]]$observed_nll),
    sapply(ok_mirt, function(m) model_states[[paste0("mirt_", m)]]$observed_nll)
  )
)

perf <- merge(metric_rows, elapsed, by = "method", all.x = TRUE)
setcolorder(perf, c("method", "elapsed_sec", "observed_nll", "ability_rmse", "loading_rmse_free", "loading_rmse_full", "corr_rmse"))
perf[order(method)]
```

```{r mirt-method-timings}
mirt_fit_status[order(status, elapsed_sec)]
```

```{r all-approach-timings}
timing_all <- rbindlist(list(
  bigirt_fit_status[, .(method, status, elapsed_sec, error_message)],
  mirt_fit_status[, .(method = paste0("mirt_", method), status, elapsed_sec, error_message)]
), use.names = TRUE, fill = TRUE)

timing_all[order(status, elapsed_sec)]
```

## 6) Recovery plots

### Ability recovery by factor

```{r ability-by-factor}
ability_dt <- person_norm[source != "truth", .(
  id = person,
  factor,
  method = source,
  true = referenceVal,
  est = value
)]

ability_dt <- compress_scatter(ability_dt, "true", "est", p = 0.98)

ggplot(ability_dt, aes(x = x_plot, y = y_plot)) +
  geom_point(aes(shape = outlier_flag, color = outlier_flag), alpha = 0.30, size = 1.0) +
  geom_abline(intercept = 0, slope = 1, linetype = 2, color = "red") +
  facet_grid(factor ~ method) +
  scale_shape_manual(values = c(in_range = 16, clipped_outlier = 17)) +
  scale_color_manual(values = c(in_range = "grey35", clipped_outlier = "darkorange3")) +
  theme_bw() +
  labs(
    title = "Ability recovery by factor (normalized scale, 98% clipping)",
    x = "Truth (clipped at 98% |value|)",
    y = "Estimate (clipped at 98% |value|)",
    shape = NULL,
    color = NULL
  )
```

### Loading recovery by factor (free entries)

```{r loading-by-factor}
loading_dt <- loading_free[, .(
  method = source,
  factor,
  true = referenceVal,
  est = value,
  abs_err = abs(value - referenceVal)
)]

loading_dt <- compress_scatter(loading_dt, "true", "est", p = 0.98)

ggplot(loading_dt, aes(x = x_plot, y = y_plot)) +
  geom_point(aes(shape = outlier_flag, color = outlier_flag), alpha = 0.35, size = 1.1) +
  geom_abline(intercept = 0, slope = 1, linetype = 2, color = "red") +
  facet_grid(factor ~ method) +
  scale_shape_manual(values = c(in_range = 16, clipped_outlier = 17)) +
  scale_color_manual(values = c(in_range = "grey35", clipped_outlier = "darkorange3")) +
  theme_bw() +
  labs(
    title = "Loading recovery on free entries (normalized scale, 98% clipping)",
    x = "Truth (clipped at 98% |value|)",
    y = "Estimate (clipped at 98% |value|)",
    shape = NULL,
    color = NULL
  )
```

### Per-factor ability correlation

```{r factor-correlation}
ability_corr_dt <- ability_dt[, .(corr = cor(true, est)), by = .(method, factor)]

ggplot(ability_corr_dt, aes(x = factor, y = corr, fill = method)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.65) +
  coord_cartesian(ylim = c(NA, 1)) +
  theme_bw() +
  labs(
    title = "Per-factor ability correlation with truth",
    x = "Factor",
    y = "Correlation"
  )

ability_corr_dt[order(factor, -corr)]
```

### Per-factor loading correlation

```{r loading-correlation}
loading_corr_dt <- loading_dt[, .(corr = cor(true, est)), by = .(method, factor)]

ggplot(loading_corr_dt, aes(x = factor, y = corr, fill = method)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.65) +
  coord_cartesian(ylim = c(0, 1)) +
  theme_bw() +
  labs(
    title = "Per-factor loading correlation with truth (free entries)",
    x = "Factor",
    y = "Correlation"
  )
```

### Ability-correlation recovery

```{r ability-corr-recovery}
corr_plot_dt <- corr_raw[source != "truth" & row < col, .(
  method = source,
  row,
  col,
  true = referenceVal,
  est = value
)]

ggplot(corr_plot_dt, aes(x = true, y = est)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_abline(intercept = 0, slope = 1, linetype = 2, color = "red") +
  facet_wrap(~ method) +
  coord_equal() +
  theme_bw() +
  labs(
    title = "Estimated latent correlations vs truth",
    x = "True correlation",
    y = "Estimated correlation"
  )
```

### Loading absolute-error distribution

```{r loading-error-distribution}
err_cap <- as.numeric(quantile(loading_dt$abs_err, probs = 0.98, na.rm = TRUE))
if(!is.finite(err_cap) || err_cap <= 0) err_cap <- max(loading_dt$abs_err, na.rm = TRUE)
loading_dt[, abs_err_plot := pmin(abs_err, err_cap)]
loading_dt[, err_outlier := fifelse(abs_err > err_cap, "clipped_outlier", "in_range")]

ggplot(loading_dt, aes(x = method, y = abs_err_plot, fill = method)) +
  geom_violin(trim = TRUE, alpha = 0.45, color = NA) +
  geom_boxplot(width = 0.15, outlier.shape = NA) +
  geom_jitter(
    aes(color = err_outlier, shape = err_outlier),
    width = 0.12,
    height = 0,
    alpha = 0.35,
    size = 1.0
  ) +
  facet_wrap(~ factor, nrow = 1) +
  scale_shape_manual(values = c(in_range = 16, clipped_outlier = 17)) +
  scale_color_manual(values = c(in_range = "grey35", clipped_outlier = "darkorange3")) +
  theme_bw() +
  labs(
    title = "Distribution of |loading error| on free entries (98% clipping)",
    x = NULL,
    y = "Absolute error (clipped at 98% quantile)",
    color = NULL,
    shape = NULL
  )
```

### Probability calibration on observed responses

```{r calibration}
calib <- rbindlist(lapply(names(model_states), function(m){
  data.table(method = m, p = model_states[[m]]$pcorrect, y = sim$dat$score)
}))

calib[, bin := cut(p, breaks = seq(0, 1, by = 0.1), include.lowest = TRUE)]
calib_dt <- calib[, .(
  mean_pred = mean(p),
  obs_rate = mean(y),
  n = .N
), by = .(method, bin)]

ggplot(calib_dt, aes(x = mean_pred, y = obs_rate, color = method)) +
  geom_abline(intercept = 0, slope = 1, linetype = 2, color = "grey50") +
  geom_point(aes(size = n), alpha = 0.85) +
  geom_line(alpha = 0.6) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  theme_bw() +
  labs(
    title = "Probability-correct calibration",
    x = "Mean predicted probability correct",
    y = "Observed proportion correct",
    size = "Count"
  )

calib[, .(
  logloss = -mean(log(ifelse(y == 1, p, 1 - p))),
  brier = mean((y - p)^2)
), by = method][order(logloss)]
```

## 7) Loading heatmaps

```{r loading-heatmaps}
hm_dt <- loading_norm[, .(
  item = as.integer(row),
  factor = as.integer(sub("^F", "", col)),
  loading = value,
  method = fifelse(source == "truth", "True (normalized)", source)
)]

q <- quantile(hm_dt$loading, probs = c(0.025, 0.975), na.rm = TRUE)
true_rng <- range(hm_dt[method == "True (normalized)", loading], na.rm = TRUE)
lim_lo <- min(q[[1]], true_rng[[1]])
lim_hi <- max(q[[2]], true_rng[[2]])
lim <- max(abs(c(lim_lo, lim_hi)))
ggplot(hm_dt, aes(x = factor, y = item, fill = loading)) +
  geom_tile() +
  scale_fill_gradient2(
    low = "navy",
    mid = "white",
    high = "firebrick",
    midpoint = 0,
    limits = c(-lim, lim)
  ) +
  facet_wrap(~ method, nrow = 1) +
  scale_x_continuous(breaks = seq_len(Nscales)) +
  theme_bw() +
  labs(
    title = "Loading matrices (normalized comparison scale)",
    x = "Latent factor",
    y = "Item"
  )

truth_dt <- loading_norm[source == "truth" & form == "normalized", .(
  row,
  col,
  truth = value
)]

diff_dt <- merge(
  loading_norm[source != "truth" & form == "normalized", .(row, col, source, value)],
  truth_dt,
  by = c("row", "col"),
  all.x = TRUE
)[, .(
  item = as.integer(row),
  factor = as.integer(sub("^F", "", col)),
  diff = value - truth,
  method = source
)]

q_diff <- quantile(diff_dt$diff, probs = c(0.025, 0.975), na.rm = TRUE)
lim_diff <- max(abs(q_diff))

ggplot(diff_dt, aes(x = factor, y = item, fill = diff)) +
  geom_tile() +
  scale_fill_gradient2(
    low = "navy",
    mid = "white",
    high = "firebrick",
    midpoint = 0,
    limits = c(-lim_diff, lim_diff)
  ) +
  facet_wrap(~ method, nrow = 1) +
  scale_x_continuous(breaks = seq_len(Nscales)) +
  theme_bw() +
  labs(
    title = "Loading discrepancy vs truth (normalized scale)",
    x = "Latent factor",
    y = "Item",
    fill = "Estimate - truth"
  )
```

## 8) Notes

- `bigirt_methods` controls which `bigIRT` estimators run. Leave only `LaplaceDirect` to skip JML, or add `bigirt_methods$JML <- list()` to include it.
- It tests `mirt` with the methods listed in `mirt_methods` and records timing/status for each.
- Recovery tables and plots now use `compareIRTmodels()` as the package-level extraction/normalization helper, so truth, `bigIRT`, and `mirt` all flow through one comparison pipeline.
