---
title: "`r params$title`"
output:
  html_document:
    self_contained: true
    toc: true
    toc_float: true
    theme: readable
params:
  fit: NULL
  title: "bigIRT model report"
  method: "profile"
  max_icc: 9
  cores: 1
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 5.5, dpi = 110)
library(bigIRT)
fit <- params$fit
st <- fit$laplaceStatus
```

## Convergence

```{r convergence}
cc <- try(checkConvergence(fit, cores = params$cores), silent = TRUE)
if (!inherits(cc, "try-error")) print(cc) else cat("Convergence detail unavailable.\n")
```

`r if (!isTRUE(st$converged)) "**This fit did not meet the gradient tolerance.** Read everything below with that in mind: the block table above says which parameters are responsible." else ""`

## Model

```{r glance}
g <- glanceIRT(fit)
knitr::kable(t(g), col.names = "value")
```

The log-likelihood is marginal, with the person parameters integrated out, so the
degrees of freedom count item parameters and hyperparameters only.

## Item parameters

```{r items}
ti <- try(tidyIRT(fit, conf.int = TRUE, method = params$method, cores = params$cores),
          silent = TRUE)
if (!inherits(ti, "try-error")) {
  knitr::kable(head(ti[order(ti$term, ti$item), ], 60), digits = 3, row.names = FALSE)
} else cat("Item parameters unavailable.\n")
```

`r if (identical(params$method, "profile")) "Standard errors use the closed-form profile weight and run about six per cent small; the hessian method differences the analytic gradient instead and is exact up to a block-diagonal term." else "Standard errors come from differencing the analytic gradient."`

## Test information

```{r information}
plot(fit, type = "information")
```

```{r reliability}
r <- try(reliability(fit), silent = TRUE)
if (!inherits(r, "try-error")) knitr::kable(r, digits = 3, row.names = FALSE)
```

## Person and item coverage

```{r wright, fig.height=6}
plot(fit, type = "wright")
```

## Item fit

```{r itemfit}
itf <- try(itemFit(fit), silent = TRUE)
if (!inherits(itf, "try-error")) {
  itf <- itf[order(-abs(itf$outfit - 1)), ]
  knitr::kable(head(itf, 20), digits = 3, row.names = FALSE,
               caption = "Items ordered by departure of outfit from 1.")
}
```

Infit and outfit are Rasch-family statistics; their reference distributions are
only well behaved for one- and two-parameter models.

## Observed against model-implied curves

The items whose outfit departs furthest from 1. Bins are formed on estimated
ability, which is shrunk and noisy, so the observed curve is flattened slightly
relative to the model; read a one-sided or sharply localised departure as
misfit, not a mild symmetric one.

```{r icc, fig.height=7}
if (!inherits(itf, "try-error")) {
  worst <- utils::head(itf$item, params$max_icc)
  idx <- match(worst, as.data.frame(fit$itemPars)$Item)
  idx <- idx[!is.na(idx)]
  if (length(idx)) plot(fit, type = "icc", items = idx)
}
```

## Held-out performance

```{r heldout}
ho <- try(heldoutMetrics(fit), silent = TRUE)
if (!inherits(ho, "try-error") && !is.null(ho)) {
  knitr::kable(as.data.frame(ho), digits = 4, row.names = FALSE)
  cat("\nThe item-base-rate benchmark predicts each item's training success rate",
      "and ignores the person, so it is this model with zero discrimination.\n")
} else cat("No held-out responses; refit with `trainingRows` to score them.\n")
```
